---
title: "Gamma-Poisson Model of User Heterogeneity"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(rstan)

#--------------
# Color scheme
#--------------
md <- read.csv("MaterialDesign.csv", stringsAsFactors=FALSE)
md500 <- md[6,]
md700 <- md[8,]

theme_gr <- function(base_size=12, base_family="") {
    gray <- md500$grey
    theme_bw(base_size = base_size, base_family = base_family) +
        theme(
            line = element_line(color = gray),
            rect = element_rect(fill = "white", color = gray, size = 0.1),
            text = element_text(color = gray),
            axis.ticks = element_line(color = gray, size = 0.1),
            legend.key = element_rect(color = NA),
            panel.border = element_rect(color = gray),
            panel.grid = element_blank(),
            strip.background = element_rect(fill = "white", color = NA),
            strip.text.x = element_text(hjust = 0, size = 10, color = gray),
            strip.text.y = element_text(hjust = 0, size = 8, color = gray,
                                        angle = 0),
            legend.position="none",
            plot.title = element_text(hjust = 0, size=14, color = "black")
        )
}
```

## Data

```{r params, include=FALSE}
num_users <- 1000
num_observations_per_user <- 100
gamma_shape <- 1.5
gamma_rate <- 1/50
num_months <- num_observations_per_user - 1
date_range <- ymd("2017-01-01") - months(seq(0, num_months))
```

```{r generate-data, include=FALSE}
generate_poisson_data <- function(x) {
    x %>%
        rpois(n = num_observations_per_user, lambda = .) %>%
        data_frame(value = ., lambda = x, t = date_range)
}

lambda_vec <- rgamma(n = num_users, shape = gamma_shape, rate = gamma_rate) 

df <- lambda_vec %>%
    sort() %>%
    map_df(generate_poisson_data, .id = 'draw') %>%
    mutate(draw = as.integer(draw))
```

```{r, include=FALSE}
df2 <- df %>%
    spread(t, value)

df3 <- df2 %>%
    select(-lambda, -draw) %>%
    as.data.frame()

Y <- df2 %>%
    select(-lambda, -draw) %>%
    as.matrix()
```

In this example, suppose we have data on the number of times per month that users of amazingcatnews.com visit the site. We have a 1000 x 100 data frame. The rows are users and the columns are months: row 1 represents the number of site visits per month between Oct 2008 and Jan 2017 for user 1, row 2 corresponds to visits per month for user 2, etc.

```{r, echo=FALSE}
df3[1:10, 1:6]
```

Plotting the data below, most users are relatively inactive, but there's a long tail of very active users. Most of the user time traces are plotted in gray, but a few examples are highlighted in teal.

```{r, echo=FALSE, fig.width=12, fig.height=6}
p <- ggplot() +
    geom_line(data = df, aes(t, value, group = draw),
              size = 0.5, alpha = 0.1, color = md500$grey) +
    geom_line(data = df %>% filter(draw %in% c(200, 600, 800, 975, 990)), aes(t, value, group = draw),
              size = 0.5, alpha = 1.0, color = md700$teal) +
    labs(x = '', y = '', title = '') +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %Y",
                 expand = c(0.0, 0)) +
    scale_y_continuous(breaks = seq(0, 500, 50)) +
    theme_gr() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggExtra::ggMarginal(p, data = df, type = "histogram", margins = "y")
```


## Fitting model to data using RStan

We can use Stan to fit a model to the data.

### Model

User $i$ visited amazingcatnews.com $Y_{i,j}$ times in month $j$. The number of monthly visits is Poisson-distributed with rate parameter $\lambda_i$. The rate parameters are sampled from a gamma distribution with scale parameter $\alpha$ and rate parameter $\beta$. Flat priors are assigned to $\alpha$ and $\beta$.

$$Y_{i,j} \sim \text{Poisson}(\lambda_i)$$
$$\lambda_i \sim \text{Gamma}(\alpha, \beta)$$
$$\alpha \sim \text{Uniform}(0, 10)$$
$$\beta \sim \text{Uniform}(0, 1)$$

###Define Stan model

```{stan output.var="pg_model"}
data {
    int<lower=1> N; // num users
    int<lower=1> K; // num observations per user
    int<lower=0> Y[N,K];
}
parameters {
    real<lower=0> alpha; // gamma scale
    real<lower=0> beta;  // gamma shape
    real<lower=0> lambda[N]; // poisson rate by user
}
model {
    for(i in 1:N) {
        lambda[i] ~ gamma(alpha,beta);
        Y[i] ~ poisson(lambda[i]);
    }
}
```

### Stan fit

```{r, eval=F}
data = list(N = num_customers, K = num_observations_per_user,
            Y = Y)

fit <- stan(file = 'poisson_gamma.stan', data = data,
            iter = 1500, warmup = 500, chains = 1)
```

```{r, echo=F}
load("stan_fit.RData")
print(fit, digits_summary = 3, pars=c('alpha', 'beta',
                                      'lambda[1]', 'lambda[2]',
                                      'lambda[3]', 'lambda[4]'))
```

The true values used to generate the data were $\alpha = 1.5$ and $\beta = 0.02$.
